\textbf{Betelwise1: A Hybrid Conventional-CNN Approach for Efficient Planetary Seismic Event Detection and Analysis}

### 2025-05-24 paper.tex
adding background

\section{Introduction}
The exploration of planetary bodies beyond Earth, such as the Moon and Mars, 
provides invaluable insights into planetary formation, evolution, and internal 
structure \cite{b8}. Seismology plays a pivotal role in these endeavors, as seismic waves 
carry information about the interior of a planet, including its crustal thickness, mantle 
structure, and core state \cite{b9, b10}. Missions like the Apollo Lunar Surface 
Experiments Package (ALSEP) \cite{b11} and the recent NASA Mars InSight mission \cite{b12} 
have deployed seismometers that generate vast quantities of data. However, transmitting 
this continuous stream of seismic data back to Earth is a formidable challenge due to 
severe constraints on bandwidth, power, and data volume inherent in deep-space missions 
\cite{b13}.

Traditional seismic event detection algorithms, often relying on methods like the Short-Term Average 
to Long-Term Average (STA/LTA) ratio \cite{b14}, are computationally inexpensive but tend to produce 
a high number of false positives, especially in noisy extra-terrestrial environments \cite{b15}. This 
inefficiency leads to the transmission of large volumes of non-pertinent data, consuming precious mission 
resources. Consequently, there is a critical need for robust, efficient, and autonomous onboard systems 
capable of accurately identifying true seismic events and prioritizing data for telemetry.

Machine learning (ML), particularly deep learning techniques like Convolutional Neural Networks (CNNs), 
has emerged as a powerful tool for analyzing complex patterns in geophysical data \cite{b16, b17}. In 
planetary seismology, CNNs have shown promise for tasks such as detecting moonquakes from Apollo data. 
For instance, Civilini et al. (2021) demonstrated the use of CNNs trained on spectrograms of Earth-based 
seismic events, employing transfer learning to successfully identify moonquakes in the Apollo PSE and 
LSPE datasets \cite{favPaper}. Their work highlighted the potential of ML to overcome the limitations of local 
data scarcity by leveraging non-local training sets.

Building upon these advancements, this paper introduces "Asteroseismology," a novel AI-driven framework
designed for efficient and accurate seismic event detection on extra-terrestrial missions. Our 
approach uniquely integrates established classical seismic processing techniques with a lightweight CNN. 
This hybrid system aims to harness the interpretability and efficiency of conventional algorithms for 
initial event candidacy while leveraging the pattern recognition capabilities of CNNs for final,
high-precision event verification. The Asteroseismology framework is designed for onboard implementation, 
emphasizing computational efficiency and adaptability to diverse planetary environments through 
tunable parameters. By processing data from both lunar (ALSEP) and Martian (InSight) missions, we 
demonstrate the system's potential to significantly reduce false positives, optimize data transmission, 
and enhance the scientific return of current and future planetary seismology missions.

The main contributions of this paper are:
\begin{itemize}
    \item A hybrid seismic event detection framework combining classical phase-picking with a CNN classifier for enhanced accuracy and efficiency.
    \item A detailed methodology for data preprocessing, initial event detection, and CNN-based refinement tailored for planetary seismic data.
    \item Demonstration of the system's computational efficiency and adaptability using data from Apollo and InSight missions.
    \item Initial performance results indicating high detection accuracy with a low false positive rate, paving the way for autonomous onboard seismic analysis.
\end{itemize}

This paper is organized as follows: Section \ref{sec:related_work} reviews related work in seismic event 
detection and machine learning applications in seismology. Section \ref{sec:methodology} details the 
proposed Asteroseismology framework. Section \ref{sec:results_discussion} presents initial results and 
discusses their implications. Finally, Section \ref{sec:conclusion} concludes the paper and outlines future
research directions.

@@

\bibitem{favPaper} F. Civilini, R.C. Weber, Z. Jiang, D. Phillips, and W. David Pan, ``Detecting moonquakes using convolutional neural networks, a non-local training set, and transfer learning,'' \emph{Geophys. J. Int.}, vol. 225, no. 3, pp. 2120--2134, Mar. 2021.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73. 
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished. 
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press. 
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982]. 
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\bibitem{b8} S. C. Solomon et al., ``New perspectives on ancient Mars,'' \emph{Science}, vol. 307, no. 5713, pp. 1214--1220, Feb. 2005.
\bibitem{b9} P. Lognonné, ``Planetary seismology,'' \emph{Annu. Rev. Earth Planet. Sci.}, vol. 33, pp. 571--604, May 2005.
\bibitem{b10} W. B. Banerdt et al., ``Initial results from the InSight mission on Mars,'' \emph{Nature Geoscience}, vol. 13, pp. 183--189, Mar. 2020.
\bibitem{b11} Y. Nakamura, G. V. Latham, and H. J. Dorman, ``Apollo lunar seismic experiment – Final summary,'' \emph{J. Geophys. Res.}, vol. 87, no. S01, p. A117, 1982.
\bibitem{b12} P. Lognonné et al., ``SEIS: Insight’s seismic experiment for internal structure of Mars,'' \emph{Space Sci. Rev.}, vol. 215, no. 12, Feb. 2019.
\bibitem{b13} R. D. Lorenz, ``Energy cost of acquiring and transmitting science data on deep-space missions,'' \emph{J. Spacecr. Rockets}, vol. 52, no. 6, pp. 1693--1697, Nov. 2015.
\bibitem{b14} R. Allen, ``Automatic earthquake recognition and timing from single traces,'' \emph{Bull. Seismol. Soc. Am.}, vol. 68, no. 5, pp. 1521--1532, Oct. 1978.
\bibitem{b15} M. P. Panning et al., ``Expected seismicity and the seismic noise environment of Europa,'' \emph{J. Geophys. Res. Planets}, vol. 123, no. 1, pp. 163--179, Jan. 2018.
\bibitem{b16} T. Perol, M. Gharbi, and M. Denolle, ``Convolutional neural network for earthquake detection and location,'' \emph{Sci. Adv.}, vol. 4, no. 2, p. e1700578, Feb. 2018.
\bibitem{b17} Z. E. Ross, M. J. Meier, and E. Hauksson, ``Generalized seismic phase detection with deep learning,'' \emph{Bull. Seismol. Soc. Am.}, vol. 108, no. 5A, pp. 2894--2901, Oct. 2018.
\bibitem{b18} G. C. Beroza and S. G. Shelly, ``Detecting and characterizing earthquake swarms,'' \emph{Science}, vol. 319, no. 5865, pp. 906--909, Feb. 2008. (Note: this is a review mentioning template matching, original template matching papers are older, e.g., Gibbons \& Ringdal 2006)
\bibitem{b19} S. Mostafa, M. R. Zare, and S. M. J. Aleali, ``Seismic signal classification using support vector machine and particle swarm optimization,'' \emph{J. Seismol.}, vol. 19, pp. 619--630, 2015.
\bibitem{b20} W. J. N. R. Mousavi, S. M., Horton, S. P., et al., ``CRED: A deep residual network of convolutional and recurrent units for earthquake signal detection,'' \emph{Sci. Rep.}, vol. 9, no. 1, p. 10267, Jul. 2019.
\bibitem{b21} C. Clinton et al., ``The Marsquake catalogue from InSight, S01 to S08,'' \emph{Phys. Earth Planet. Inter.}, vol. 310, p. 106595, Jan. 2021.
\bibitem{b22} A. Odena, C. Olah, and J. Shlens, ``Conditional image synthesis with auxiliary classifier GANs,'' in \emph{Proc. Int. Conf. Mach. Learn. (ICML)}, 2017, pp. 2642--2651.
\bibitem{b26} F. Baer, M., \& Kradolfer, U. (1987). An automatic phase picker for local and teleseismic events. \emph{Bulletin of the Seismological Society of America}, 77(4), 1437-1445.

>>>>

\section{Introduction}
    The seismic exploration of planetary bodies like the Moon and Mars is fundamental to understanding their internal
    structure, geological evolution, and potential for past or present habitability \cite{Lognonne2005, Lognonne2019}.
    Data from missions such as the Apollo Passive Seismic Experiment (PSE) \cite{Nakamura1982} and the ongoing Mars
    InSight mission with its SEIS instrument\cite{Lognonne2019} have provided invaluable windows into these worlds. A
    primary constraint in planetary seismology, as emphasized by \cite{Civilini2021}, is the costly power and bandwidth
    requirements associated with delivering continuous seismic data back to Earth from distant missions.

    A fundamental challenge in planetary seismology is the efficient management and transmission of the vast quantities
    of continuous seismic data generated by lander-based instruments. Deep-space missions operate under severe
    constraints on power, bandwidth, and data volume (Lorenz, 2015, as cited in Civilini et al., 2021; Lognonné et al.,
    2019). For instance, the SEIS instrument on the Mars InSight mission, while typically downlinking continuous data at
    2 samples per second (sps), generates a significant data volume; \cite{Lognonne2019} note that its full data
    generation capability is orders of magnitude larger than its telemetry allocation. \cite{Civilini2021} quantified
    that the telemetry for SEIS consumes approximately 1.5\% of the InSight lander's total power output. If we were to
    consider data rates comparable to those from some Apollo experiments, which utilized sampling rates such as 6.625 Hz
    for the Passive Seismic Experiment (PSE) mid-period instruments \cite{Nunn2020}, the data volume and consequently
    the power required for transmission by a modern system like SEIS would consume an even more substantial portion of a
    lander's available power. Extrapolating further, \cite{Civilini2021} projected that a similar seismic setup for a
    mission to Europa could demand 8.8–13.2\% of the lander's power solely for seismic data transmission. These figures
    underscore the critical necessity for sophisticated on-board data processing and event detection algorithms to
    prioritize scientifically valuable data segments for downlink, thereby making the most efficient use of limited
    mission resources.

    Traditional automatic event detection algorithms, like the Short-Term Average to Long-Term Average (STA/LTA) ratio
    (Allen, 1982, as cited in \cite{Civilini2021}), while computationally inexpensive, often struggle with the low
    signal-to-noise ratios and complex noise environments encountered on planetary surfaces. This can lead to a high
    rate of false positives, thereby inefficiently utilizing the scarce telemetry resources \cite{Civilini2021}.
    Furthermore, the unique characteristics of seismic wave propagation on bodies like the Moon, with its highly
    scattering near-surface layer leading to prolonged codas \cite{Dainty1981, Nakamura1982} , make robust event
    detection and phase-picking particularly challenging.

    In response to these challenges, machine learning (ML) approaches, especially deep learning techniques like
    Convolutional Neural Networks (CNNs), have gained traction in planetary science. \cite{Civilini2021} effectively
    demonstrated that CNNs, trained on spectrograms of Earth-based seismic events and utilizing transfer learning, could
    accurately catalog moonquakes from Apollo PSE and LSPE data, even in the absence of extensive local training
    datasets. This work underscored the potential of ML to generalize from available data and perform robust
    classification.

    This paper introduces "Asteroseismology," an AI-enhanced seismic event detection framework specifically designed to
    address the data optimization needs of extra-terrestrial missions. Our framework uniquely integrates a pipeline of
    classical phase-picking algorithms with a lightweight CNN. This hybrid approach begins with conventional signal
    conditioning, including automatic bandpass filtering, outlier removal, and normalization. Candidate seismic events
    are then identified using STA/LTA analysis, followed by a novel filtering stage based on characteristic function
    slopes to refine these initial picks. The core innovation lies in then feeding these refined candidate events,
    represented as 1D waveform segments augmented with auxiliary statistical features, into a CNN. This CNN, trained on
    a diverse dataset including lunar data from the Apollo missions and Martian data from the InSight mission, is tasked
    with the final discrimination between true seismic events and noise. By leveraging the efficiency of classical
    algorithms for initial data reduction and candidate selection, our system aims to reduce the computational burden on
    the CNN, allowing it to perform high-precision classification. Asteroseismology is designed for computational
    efficiency and adaptability through tunable parameters, making it a strong candidate for future onboard autonomous
    data processing, thereby significantly enhancing the scientific return of planetary seismology missions by
    optimizing precious data transmission resources.

\section{Background}
    \subsection{Lunar Seismology}
        The foundation of extra-terrestrial seismology was laid by the NASA Apollo missions (1969-1977). The Apollo
        Lunar Surface Experiments Package (ALSEP) successfully deployed a network of four primary seismic stations
        (Apollo 12, 14, 15, and 16), complemented by the short-lived Apollo 11 seismometer and the Lunar Seismic
        Profiling Experiment (LSPE) geophone array on Apollo 17 \cite{Nakamura1982, Lognonne2005, Nunn2020}
        %(Nakamura et al., 1981; Lognonné, 2005; Nunn et al., 2020a).
        These Passive Seismic Experiments (PSE) typically included three-component long-period (LP, later referred to as
        mid-period or MP) seismometers and a single-component short-period (SP) seismometer
        \cite{Nunn2020}%(Nunn et al., 2020a).

        Lunar seismic data revealed a surprisingly active Moon, characterized by several types of moonquakes: deep
        moonquakes (DMQs) occurring at depths of 700-1200 km and exhibiting strong tidal periodicities; shallow
        moonquakes (or HFTs - High-Frequency Teleseismic events) at shallower depths (50-220 km); thermal moonquakes
        caused by surface temperature variations; and meteoroid impacts \cite{Nakamura1982, Lognonne2005}
        %(Nakamura et al., 1982; Lognonné, 2005)
        . A key characteristic of lunar seismograms is the intense scattering within the lunar megaregolith, resulting
        in prolonged codas and making phase identification, particularly for S-waves, exceptionally difficult
        \cite{Dainty1981, Garcia2019}
        %(Dainty and Toksöz, 1981; Garcia et al., 2019).

        The initial cataloging of these events, notably by \cite{Nakamura1982}, relied heavily on visual inspection and
        waveform cross-correlation to identify repeating DMQs from distinct source regions or "nests." These efforts
        provided the first constraints on the Moon's internal structure, including its crust, mantle, and the likely
        presence of a small, partially molten core \cite{Weber2011,Garcia2019}
        %(Weber et al., 2011; Garcia et al., 2019).

        More recently, with the digitization and improved accessibility of Apollo data, new analytical techniques have
        been applied. Traditional methods continue to be refined, for example, for re-evaluating Apollo 17 LSPE data
        (Heffels et al., 2017). Machine learning has also begun to make inroads. Knapmeyer-Endrun \& Hammer (2015)
        utilized Hidden Markov Models for event detection in Apollo 16 data. Significantly, Civilini et al. (2021)
        demonstrated the successful application of Convolutional Neural Networks (CNNs), trained on spectrograms of
        Earth-based seismic events via transfer learning, to detect and catalog moonquakes from both PSE and LSPE data.
        This approach highlighted the potential of ML to identify events even with limited or non-local training data,
        by learning characteristic spectral features. Majstorović et al. (2024, preprint/submitted) further explored ML
        by using random forests to classify DMQ source regions based on lunar orbital parameters related to their
        occurrence times, showcasing an approach independent of waveform data for initial source region estimation.

    \subsection{ Martian Seismology}
    Mars has been the focus of renewed seismic interest. The first attempt to place a seismometer on Mars was with the
    Viking landers in 1976. The Viking 2 lander's seismometer, mounted on the lander deck, provided data primarily on
    wind-induced lander vibrations, and only one candidate seismic event was tentatively identified (Anderson et al.,
    1977; Lognonné, 2005). The on-deck deployment and instrument sensitivity were major limitations.

@@

    \bibitem{Civilini2021} F. Civilini, R.C. Weber, Z. Jiang, D. Phillips, and W. David Pan, ``Detecting moonquakes using convolutional neural networks, a non-local training set, and transfer learning,'' \emph{Geophys. J. Int.}, vol. 225, no. 3, pp. 2120--2134, Mar. 2021.
    \bibitem{Lognonne2005}  P. Lognonné, ``Planetary seismology,'' \emph{Annu. Rev. Earth Planet. Sci.}, vol. 33, pp. 571--604, May 2005.
    \bibitem{Lognonne2019} P. Lognonné, W. B. Banerdt, V. Dehant, et al., ``SEIS: The seismic experiment for internal structure of InSight,'' \emph{Space Sci. Rev.}, vol. 215, Art. no. 12, Feb. 2019, doi: 10.1007/s11214-018-0574-6.
    \bibitem{Nakamura1982} Y. Nakamura, G. V. Latham, and H. J. Dorman, ``Apollo lunar seismic experiment – Final summary,'' \emph{J. Geophys. Res.}, vol. 87, no. S01, p. A117, 1982.
    \bibitem{Nakamura1981} Y. Nakamura, ``Lunar seismicity and the internal structure of the Moon,'' \emph{J. Geophys. Res.}, vol. 86, no. B10, pp. 9413--9424, Oct. 1981.
    \bibitem{Lorenz2015} R. D. Lorenz, ``Energy cost of acquiring and transmitting science data on deep-space missions,'' \emph{J. Spacecr. Rockets}, vol. 52, no. 6, pp. 1693--1697, Nov. 2015.
    \bibitem{Nunn2020} C. Nunn, R. F. Garcia, Y. Nakamura, et al., ``Lunar seismology: A data and instrumentation review,'' \emph{Space Sci. Rev.}, vol. 216, no. 89, pp. 1--37, 2020.
    \bibitem{Dainty1981} A. Dainty and M. Toksöz, ``Seismic codas on the Earth and the Moon: A comparison,'' \emph{Phys. Earth Planet. Inter.}, vol. 26, no. 4, pp. 250--260, 1981.
    \bibitem{Garcia2019} R. F. Garcia, A. Khan, M. Drilleau, et al., ``Lunar seismology: An update on interior structure models,'' \emph{Space Sci. Rev.}, vol. 215, no. 50, pp. 1--50, 2019.
    \bibitem{Weber2011} R. C. Weber, P. Y. Lin, E. J. Garnero, Q. Williams, and P. Lognonné, ``Seismic detection of the lunar core,'' \emph{Science}, vol. 331, no. 6015, pp. 309--312, Jan. 2011.
    % \bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73. 
    % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 
    % \bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished. 
    % \bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press. 
    % \bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982]. 
    % \bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
    % \bibitem{b8} S. C. Solomon et al., ``New perspectives on ancient Mars,'' \emph{Science}, vol. 307, no. 5713, pp. 1214--1220, Feb. 2005.
    % \bibitem{b9} P. Lognonné, ``Planetary seismology,'' \emph{Annu. Rev. Earth Planet. Sci.}, vol. 33, pp. 571--604, May 2005.
    % \bibitem{b10} W. B. Banerdt et al., ``Initial results from the InSight mission on Mars,'' \emph{Nature Geoscience}, vol. 13, pp. 183--189, Mar. 2020.
    % \bibitem{b12} P. Lognonné et al., ``SEIS: Insight’s seismic experiment for internal structure of Mars,'' \emph{Space Sci. Rev.}, vol. 215, no. 12, Feb. 2019.
    % \bibitem{b13} R. D. Lorenz, ``Energy cost of acquiring and transmitting science data on deep-space missions,'' \emph{J. Spacecr. Rockets}, vol. 52, no. 6, pp. 1693--1697, Nov. 2015.
    % \bibitem{b14} R. Allen, ``Automatic earthquake recognition and timing from single traces,'' \emph{Bull. Seismol. Soc. Am.}, vol. 68, no. 5, pp. 1521--1532, Oct. 1978.
    % \bibitem{b15} M. P. Panning et al., ``Expected seismicity and the seismic noise environment of Europa,'' \emph{J. Geophys. Res. Planets}, vol. 123, no. 1, pp. 163--179, Jan. 2018.
    % \bibitem{b16} T. Perol, M. Gharbi, and M. Denolle, ``Convolutional neural network for earthquake detection and location,'' \emph{Sci. Adv.}, vol. 4, no. 2, p. e1700578, Feb. 2018.
    % \bibitem{b17} Z. E. Ross, M. J. Meier, and E. Hauksson, ``Generalized seismic phase detection with deep learning,'' \emph{Bull. Seismol. Soc. Am.}, vol. 108, no. 5A, pp. 2894--2901, Oct. 2018.
    % \bibitem{b18} G. C. Beroza and S. G. Shelly, ``Detecting and characterizing earthquake swarms,'' \emph{Science}, vol. 319, no. 5865, pp. 906--909, Feb. 2008. (Note: this is a review mentioning template matching, original template matching papers are older, e.g., Gibbons \& Ringdal 2006)
    % \bibitem{b19} S. Mostafa, M. R. Zare, and S. M. J. Aleali, ``Seismic signal classification using support vector machine and particle swarm optimization,'' \emph{J. Seismol.}, vol. 19, pp. 619--630, 2015.
    % \bibitem{b20} W. J. N. R. Mousavi, S. M., Horton, S. P., et al., ``CRED: A deep residual network of convolutional and recurrent units for earthquake signal detection,'' \emph{Sci. Rep.}, vol. 9, no. 1, p. 10267, Jul. 2019.
    % \bibitem{b21} C. Clinton et al., ``The Marsquake catalogue from InSight, S01 to S08,'' \emph{Phys. Earth Planet. Inter.}, vol. 310, p. 106595, Jan. 2021.
    % \bibitem{b22} A. Odena, C. Olah, and J. Shlens, ``Conditional image synthesis with auxiliary classifier GANs,'' in \emph{Proc. Int. Conf. Mach. Learn. (ICML)}, 2017, pp. 2642--2651.
    % \bibitem{b26} F. Baer, M., \& Kradolfer, U. (1987). An automatic phase picker for local and teleseismic events. \emph{Bulletin of the Seismological Society of America}, 77(4), 1437-1445.
